---
title: "Live Session Unit 11 Assignment"
author: "Scott Payne"
date: "November 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Packages required:
# install.packages("fpp")


Codetoloaddata:
```{r message=FALSE}
  library(fpp) 
  #fpp package must be installed first
  data(ukcars)
```
Plot the time series. Can you identify seasonal fluctuations and/or a trend?
<br>
There is a seasonal trend of peaks and valleys at a constant rate.
```{r}
plot(ukcars)
```
<br>
Obtain classical decomposition
```{r}
fitd <- decompose(ukcars) 
plot(fitd)
```
<br>
Do the results support the graphical interpretation from part (a)? 

<br> Yes <br>

Compute and plot the seasonally adjusted data
```{r}
ukadj <- seasadj(fitd)
plot(ukadj)
```

Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?<br>
The outlier is clearly noticeable on the trend line because it does not fit with the rest of the time series, but it does not appear to have an effect on the rest of the adjusted data.<br>
Does it make any difference if the outlier is near the end rather than in the middle of the time series? Adding extreme outliers does not significantly change the adjusted trend line, but it is very easy to spot the outlier data because it sticks out obviously from the rest of the time series data. 
```{r}
ukcars2 <- ts(c(ukcars[1:54],ukcars[55]+500,ukcars[56:113] ),start=c(1977,1),frequency=12)
fitd <- decompose(ukcars2)
uk2adj <- seasadj(fitd)
plot(uk2adj)
ukcars2 <- ts(c(ukcars[1:110],ukcars[111]+500,ukcars[112:113] ),start=c(1977,1),frequency=12)
fitd <- decompose(ukcars2)
uk2adj <- seasadj(fitd)
plot(uk2adj)
```


Use STL to decompose the series. 
```{r}
fit <- stl(ukcars, s.window=5)
plot(fit)
plot(ukcars, col="gray",
  main="UK passenger vehicle production",
  ylab="Thousands of cars", xlab="")
lines(fit$time.series[,2],col="red"
,ylab="Trend")
```
